---
title: "Danish Kings Homework Week 10"
author: "Sarah Astrup"
date: "7 March 2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, echo = FALSE}
library(tidyverse)
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the

separator: "," 

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
# FILL IN THE CODE BELOW and review the outputs
kings1 <- read.csv("data/kings.csv", na="NULL")
glimpse(kings1)

kings2 <- read_csv("data/kings.csv", na="NULL")
glimpse(kings2)

kings3 <- read.csv2("data/kings.csv",  na="NULL")
glimpse(kings3)

kings4 <- read_csv2( "data/kings.csv",  na="NULL")
glimpse(kings4)

```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object

Tidyverse is kings2 and kings4 which is downloaded into R. Kings1 and Kings3 are the original R or builded in R. 
CSV2 is used when the dataset is seperated by ";" and CSV is used when datasets are seperated by ","

2. What is the result of running `class()` on the `kings` object created with a tidyverse function.

When running class on kings 2 and kings 4 it shows: [1] "spec_tbl_df" "tbl_df"      "tbl"         "data.frame, which describes a tibble 

3. How many columns does the object have when created with these different functions?

The functions kings3 and kings4 has 1 column. 
The functions kings1 and kings2 have 11 columns. 

4. Show the dataset so that we can see how R interprets each column

```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

kings <- read_csv("data/kings.csv", na="NULL")

class(kings)

ncol(kings)

glimpse(kings)

```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r}
kings_duration <- kings %>% 
  mutate(duration = End_year-Start_year) %>% 
  mutate(midyear = Start_year + (End_year-Start_year)/2) %>% 
  na.omit(duration)

# We use the mutate() function and na.omit to remove the missing values 

glimpse(kings_duration)

```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r}
kings_average <- mean(kings_duration$duration, na.rm = TRUE)

kings_average
  
#the summarize function didn't work so we use mean() and make it a vector with kings_duration$duration
```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

```{r}
 duration_above_average <- kings_duration %>% 
  filter(duration>20)

glimpse(duration_above_average)

count(duration_above_average, sort = TRUE)
  
```



## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
duration_descending <- kings_duration %>% 
  arrange(desc(duration))

glimpse(duration_descending)
```
```{r}
#Calculating the longest reign
longest_reign <- duration_descending %>% 
  slice(c(1,2,3))

longest_reign
```

```{r}
#Calculating days 
reign_days <- longest_reign %>% 
  mutate(days=duration*365)

glimpse(reign_days)
```

## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
# midyear is mutated in colon 83 
kings_duration %>% 
  ggplot(aes(x=midyear, y=duration))+
  geom_point()+
  geom_smooth()+
  theme_bw()+
  ggtitle("Kings duration of reign through time")
```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!


